Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2021-06-22 SIR_manuscript.tex
revisions to Methods

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Sec.~\ref{app:#1}}
\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand\pder[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}

@@

We developed a deterministic model which groups individuals based on disease status and testing status. Disease states include Susceptible, Infectious and Recovered; testing states include \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive} (\fref{fig:flowchart}). The testing status of an individual in  a given disease compartment $X$ (where $X \in \{S,I,R\}$) is denoted by a subscript, namely $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively.
Two `accumulator' compartments, $N$ and $P$, are included in order to collect cumulative reported negative or positive tests. The model equations \eqref{model} and details of calculation of the basic reproduction number $\Rnum$ are presented in \appref{R0}.

@@

In order to allow parameterization of the model by the total (overall) \percap testing rate, we define the weighted size of the testing pool $W = w_S S\_u + w_I I\_u + w_R R\_u$, and calculate a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W},
\end{equation}
where $\rho$ is the \percap testing intensity for the population, defined as the number of daily tests administered in a population of size $N_0$.
Thus, the \percap testing rate for compartment $X \in \{S,I,R\}$ is 
\begin{equation}
\label{F}
F_X=\sigma w_X.
\end{equation}
For a highly sensitive test, infected people typically flow through to the ``confirmed positive'' ($I\_c$, $R\_c$) compartments and are thus not considered for further testing. Over the course of the epidemic, a sufficiently large fixed testing rate as specified in \eqref{sigma} can exhaust the pool of people available for testing, leading to a singularity when too few people are left untested to support the specified rate. Although this phenomenon does not affect our analysis of $\Rnum$, it can affect model dynamics (we present an adjustment to the model that solves this problem in \appref{singularity}).

The classical SIR model assumes a well-mixed population; homogeneity of the population (i.e., all individuals are equally susceptible and equally infectious with the same recovery rate when infected); exponentially distributed duration of infection; and large population size \citep{keeling2011modeling}. In addition to these standard assumptions, our model assumes: 
\begin{enumerate}[(i)]
\item there is a single force of infection (new cases per unit time), $\Lambda$, defined as
  \begin{equation}
  \label{Lambda}
  \Lambda=\frac{\beta}{N_0} \big(I\_u + (1-\theta\_w)(I\_n+I\_p) + (1-\theta\_c)I\_c \big),
  \end{equation}
  with transmission rate $\beta$; $\theta\_w$ is the isolation efficacy (reduction of the probability of transmission) for individuals waiting for test results, while $\theta\_c$ is the isolation efficacy for individuals who have received a ``confirmed'' positive test (Table~\ref{tab:params}). Susceptible individuals who are waiting for test results experience an additional transmission factor of $1-\theta\_w$ (\fref{fig:flowchart}). 
\item confirmed-positive individuals isolate at least as effectively as those awaiting test results, i.e.
$$0\leq\theta\_w\leq\theta\_c\leq 1.$$ 
\end{enumerate}
For simplicity we assume that tests are perfectly \emph{specific} --- uninfected individuals never test positive ($p_s=0$). Thus there are no waiting-for-positive or confirmed-positive susceptible individuals, which reduces the number of model states from 12 to 10.

@@

The Disease-Free Equilibrium (DFE) for the SIR model (Eqs.~\ref{model}) is found by setting the infected compartments to 0 and solving for the unknowns. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= \frac{\omega-\rho}{\omega} N_0, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
The corresponding \percap testing rate (Eq.~\ref{F}) for the infected compartment $I$ at DFE is one of the key analysis parameters and can be simplified as 
\begin{equation}
\label{eq:fi}
\hat F_I = (\omega\rho/(\omega-\rho))w_I/w_S \quad .
\end{equation}

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ as
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} \left(1-\Delta\right), 
\end{equation}
where the term $\beta/\gamma$ is the classical basic reproduction number for a simple SIR model \citep{keeling2011modeling}, and $0 < \Delta < 1$ is the proportional reduction due to testing and isolation processes, calculated as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
(\appref{R0} gives a detailed derivation of these expressions).
This explicit formula enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (\appref{rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Delta$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$, use the linear approximation with respect to $\rho$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Delta$ and $\omega$.

@@

\jd{Why do we use a different range of $\omega$ for the two cases?}\ali{added a sentence to the end of this paragraph to address this (I hope).}
The low-testing case \fref{pan} reflects the case where $\rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 13000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$, corresponds to a simple SIR model, in these figures is 3. Note that the different range of $\omega$ for the two cases of low and high testing intensities is due to the restriction $\rho<\omega$ which is a requirement for feasible DFE (\ref{dfe}).

>>>>

\newcommand{\fref}[1]{Fig.~\ref{#1}}
\newcommand{\appref}[1]{Appendix~\ref{app:#1}}
\newcommand{\Rlogo}{R\xspace}
\newcommand{\percap}{\emph{per capita}\xspace}
\newcommand{\Rnum}{\ensuremath{\mathcal{R}_0}\xspace}
\newcommand{\covid}{COVID-19\xspace}
\newcommand{\pro}[1]{\ensuremath{\frac{\partial #1}{\partial \rho}}}
\newcommand\pder[2]{\ensuremath{\frac{\partial#1}{\partial#2}}} %\pder y{x}

@@

Our model groups individuals based on disease status (Susceptible, Infectious or Recovered) and testing status (\emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}) (\fref{fig:flowchart}).  The testing status of an individual in a given disease compartment $X$ (where $X \in \{S,I,R\}$) is denoted by a subscript, namely $X\_u$, $X\_p$, $X\_n$ and $X\_c$, for \emph{untested}, waiting-for-\emph{positive}, waiting-for-\emph{negative}, or \emph{confirmed positive}, respectively.  Two `accumulator' compartments, $N$ and $P$, are included in order to collect cumulative reported negative or positive tests. The model equations \eqref{model} and details of calculation of the basic reproduction number $\Rnum$ are presented in \appref{R0}.

@@

In order to allow parameterization of the model by the total (overall) \percap testing rate, we define the weighted size of the testing pool $W = w_S S\_u + w_I I\_u + w_R R\_u$, and calculate a scaling parameter for testing as:
\begin{equation}
\label{sigma}
\sigma = \frac{\rho N_0}{W},
\end{equation}
where $\rho$ is the \percap testing intensity for the population, defined as the number of daily tests administered in a population of size $N_0$.
\david{Why $N_0$ rather than just $N$?}
Thus, the \percap testing rate for compartment $X \in \{S,I,R\}$ is 
\begin{equation}
\label{F}
F_X=\sigma w_X.
\end{equation}
\david{$F$ isn't a natural symbol for ``testing rate''.  How about $\mathcal T_X$?}
For a highly sensitive test, infected people typically flow through to the ``confirmed positive'' ($I\_c$, $R\_c$) compartments and are thus not considered for further testing. Over the course of the epidemic, a sufficiently large fixed testing rate as specified in \eqref{sigma} can exhaust the pool of people available for testing, leading to a singularity when too few people are left untested to support the specified rate. Although this phenomenon does not affect our analysis of $\Rnum$, it can affect model dynamics (we present an adjustment to the model that solves this problem in \appref{singularity}).

The classical SIR model assumes a well-mixed population; homogeneity of the population (i.e., all individuals are equally susceptible and equally infectious with the same recovery rate when infected); exponentially distributed duration of infection; and large population size \citep{keeling2011modeling}. In addition to these standard assumptions, our model assumes: 
\begin{enumerate}[(i)]
\item there is a single force of infection (new \david{infections} per unit time \david{per susceptible}), $\Lambda$, defined as
  \begin{equation}
  \label{Lambda}
  \Lambda=\frac{\beta}{N_0} \big(I\_u + (1-\theta\_w)(I\_n+I\_p) + (1-\theta\_c)I\_c \big),
  \end{equation}
  with transmission rate $\beta$; $\theta\_w$ is the isolation efficacy (reduction of the probability of transmission) for individuals waiting for test results, while $\theta\_c$ is the isolation efficacy for individuals who have received a ``confirmed'' positive test (Table~\ref{tab:params}). Susceptible individuals who are waiting for test results experience an additional transmission reduction factor of $1-\theta\_w$ (\fref{fig:flowchart}). 
\item confirmed-positive individuals isolate at least as effectively as those awaiting test results, i.e.,
$$0\leq\theta\_w\leq\theta\_c\leq 1.$$ 
\end{enumerate}
For simplicity we assume that tests are perfectly \emph{specific} --- uninfected individuals never test positive ($p_s=0$). Thus there are no waiting-for-positive or confirmed-positive susceptible individuals, which reduces the number of model states from 12 to 10.

@@

The Disease-Free Equilibrium (DFE) for the expanded SIR model (Eqs.~\ref{model}) is found by setting the infected compartments to 0 and solving for the unknowns. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
The corresponding \percap testing rate (Eq.~\ref{F}) for the infected compartment $I$ at DFE is one of the key analysis parameters and can be simplified as 
\begin{equation}
\label{eq:fi}
\hat F_I = (\omega\rho/(\omega-\rho))w_I/w_S \quad .
\end{equation}

The basic reproduction number, $\Rnum$, was calculated by using the next-generation matrix method \citep{van2002reproduction}. We present $\Rnum$ as
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} \left(1-\Delta\right), 
\end{equation}
where the term $\beta/\gamma$ is the classical basic reproduction number for a simple SIR model \citep{keeling2011modeling}, and $0 < \Delta < 1$ is the proportional reduction \david{actually $1-\Delta$ is the proportional reduction} due to testing and isolation processes, calculated as 
\begin{equation}
  \label{eq:del4a}
  \Delta= \frac{1}{C N_0}\big(C_1 S\_u^*+(C_2(1-\theta\_w)+C \theta\_w)S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C_{\phantom 1} &= (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 &= (\omega+\gamma)(\theta\_w \gamma+\theta\_c \omega p_I) \hat F_I,\\
C_2 &= \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2p_I\hat F_I\theta\_c.
\end{align}
(\appref{R0} gives a detailed derivation of these expressions.)
This explicit formula enables us to study the effects of testing and isolation parameters on $\Rnum$ both analytically and via numerical solutions.
We are specifically interested in parameters that could be manipulated by public health policy: isolation efficacy, $\theta\_c$ and $\theta\_w$; \percap testing intensity, $\rho$; and the rate of test return, $\omega$. In particular, we look at the partial derivatives of $\Delta$ with respect to these parameters (\appref{rho}). 
We derived general expressions for these derivatives. However, we analyzed the effect of $\omega$ on $\Delta$ for the special case of low testing intensity. Specifically, by making the restriction $\rho \ll 1$, we are able to Taylor-expand $\Delta$ at $\rho=0$, use the linear approximation with respect to $\rho$ and analyze the resulting simplified derivatives to illustrate a surprising non-monotonic relationship between $\Delta$ and $\omega$.

@@

\jd{Why do we use a different range of $\omega$ for the two cases?}\ali{added a sentence to the end of this paragraph to address this (I hope).}
The low-testing case \fref{pan} reflects the case where testing intensity $\rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e., a maximum of 1.3\% of the population per day \david{If it is 4x larger than Ontario, what makes it ``realistic''?} (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for small populations undergoing focused testing, such as a sports league or university. In these figures, the implied baseline reproduction number (for the SIR model without testing) is $\Rnum=\frac{\beta}{\gamma}=3$. Note that the different range of test return rate $\omega$ for the two cases of low and high testing intensities is due to the restriction $\rho<\omega$ which is a requirement for feasible DFE (\ref{dfe}).
